A boundary integral equation approach to computing eigenvalues of the Stokes operator

### 2019-04-05 doc/draft-01-stokes/01intro.tex

The planar incompressible Stokes equations describe
creeping flows in two dimensions.
%
Let $\Omega \subset \R^2$
be a bounded domain with $C^2$ boundary.
%
The Stokes eigenvalue problem is to find
values $k^2$ such that

@@

As noted in~\cite{zhao2015robust}, integral
equation based methods provide several advantages.
%
Because the BIE is defined on the
boundary alone, there is a reduction in the
dimension of the domain to be discretized.
%
This reduces the number of unknowns over finite
element discretizations, particularly given
that finite element methods suffer from
high-frequency ``pollution'' when $k$ is large
\cite{babuska1997pollution}.

@@

With the efficiency of the approach of
\cite{zhao2015robust} for the drum problem in mind,
we develop an integral equation based method for the
Stokes eigenvalue problem.
%
This requires that a layer
potential representation of the solution
of \cref{eq:ostokes} be given and that the resulting BIE
is not invertible precisely when $k^2$ is an eigenvalue.
%
The first requirement is straightforward to
satisfy because the
well-known layer potential representations for the
modified Stokes equation~\cite{Pozrikidis1992,biros2002embedded,
  jiang2013second,ladyzhenskaya1969mathematical}
are directly applicable.
%
Proving the invertibility of the associated operators
away from the eigenvalues is a more involved task
and forms the bulk of the theoretical component
of this paper.

@@

The rest of this paper proceeds as follows.
%
In \cref{sec:prelim}, we set the notation, provide some
mathematical preliminaries, and review the
properties of single and double layer potentials
for the oscillatory Stokes equations.
%
Then, in \cref{sec:analysis}, we develop the necessary
theory for proving the main results of this work 
(\cref{thm:dlmain,thm:cfmain}),
which show that the BIEs resulting from these
layer potential representations are not invertible
precisely when $k^2$ is an eigenvalue.
%
These theoretical developments include a detailed
discussion of the uniqueness of oscillatory Stokes
boundary value problems in exterior domains.
%
To the best of our knowledge, the invertibility
and uniqueness results are new to the literature.
%
\Cref{sec:dets} then outlines how the Fredholm determinant
can be used in the oscillatory Stokes context.
%
In \cref{sec:numerical}, we describe the numerical
methods we use to discretize the BIEs and to perform
determinant calculations.
%
At the moderate frequencies considered in this
paper, we find that standard fast-direct
solvers provide a reasonably efficient determinant
evaluation.
%
That section also includes numerical experiments
which demonstrate some of the paper's analytical
claims as well as the effectiveness of the overall
framework.
%
Finally, we provide some concluding thoughts,
describe plans for future research,
and outline some open questions in
\cref{sec:conclusion}.

>>>>

The planar incompressible Stokes equations describe
creeping flows in two dimensions.
%
Let $\Omega \subset \R^2$
be a bounded domain with $C^2$ boundary denoted by $\Gamma$.
%
The Stokes eigenvalue problem is to find
values $k^2$ such that

@@

As noted in~\cite{zhao2015robust}, integral
equation based methods provide several advantages.
%
Because the BIE is defined on the
boundary alone, there is a reduction in the
dimension of the domain to be discretized.
%
This reduces the number of unknowns over finite
element discretizations, particularly given
that finite element methods suffer from
high-frequency ``pollution'' due to the 
large number of unknowns required to resolve
the computation when $k$ is large
\cite{babuska1997pollution}.

@@

With the efficiency of the approach of
\cite{zhao2015robust} for the drum problem in mind,
we develop an integral equation based method for the
Stokes eigenvalue problem.
%
This requires that a layer
potential representation of the solution
of \cref{eq:ostokes} be given and that the resulting BIE
is not invertible precisely when $k^2$ is an eigenvalue.
%
The first requirement is straightforward to
satisfy because 
well-known layer potential representations for the
modified Stokes equation~\cite{Pozrikidis1992,biros2002embedded,
  jiang2013second,ladyzhenskaya1969mathematical}
are directly applicable.
%
Proving the invertibility of the associated operators
away from the eigenvalues is a more involved task
and forms the bulk of the theoretical component
of this paper.

@@

The rest of this paper proceeds as follows.
%
In \cref{sec:prelim}, we set the notation, provide some
mathematical preliminaries, and review the
properties of single and double layer potentials
for the oscillatory Stokes equations.
%
Then, in \cref{sec:analysis}, we develop the necessary
theory for proving the main results of this work 
(\cref{thm:dlmain,thm:cfmain}),
which show that the BIEs resulting from these
layer potential representations are not invertible
precisely when $k^2$ is an eigenvalue.
%
These theoretical developments include a detailed
discussion of the uniqueness of oscillatory Stokes
boundary value problems in exterior domains.
%
To the best of our knowledge, the invertibility
and uniqueness results are new to the literature.
%
\Cref{sec:dets} then outlines how the Fredholm determinant
can be used in the oscillatory Stokes context.
%
In \cref{sec:numerical}, we describe the numerical
methods we use to discretize the BIEs and to perform
determinant calculations.
%
At the moderate frequencies considered in this
paper, we find that standard fast-direct
solvers provide a reasonably efficient determinant
evaluation.
%
We also present numerical experiments
which demonstrate some of the paper's analytical
claims as well as the effectiveness of the overall
framework.
%
Finally, we provide some concluding thoughts,
describe plans for future research,
and outline some open questions in
\cref{sec:conclusion}.

### 2019-04-05 doc/draft-01-stokes/02prelim.tex

\section{Mathematical Preliminaries}
\label{sec:prelim}
In this paper, vector-valued
quantities are denoted by bold, lower-case letters
(e.g. $\bh$), while tensor-valued quantities are bold
and upper-case (e.g. $\mathbf{T}$). 
Subscript indices of non-bold characters (e.g. $h_j$ or $T_{jkl}$)
are used to denote the entries within a vector ($\bh$) or tensor ($\bT$).
We use the standard Einstein summation convention; in other words, 
there is an implied sum taken over the repeated indices of 
any term (e.g. the symbol $a_{j} b_{j}$ is used to represent the sum
$\sum_{j} a_{j} b_{j}$).
If $\xx = (x_1,x_2)^\intercal$, then $\xx^\bot = (-x_2,x_1)^\intercal$.
Similarly, $\nabla^\bot = (-\partial_{x_2},\partial_{x_1})^\intercal$.
Upper-case script characters (e.g. $\mathcal{K}$) are reserved for
operators on Banach spaces, with $\mathcal{I}$ denoting the
identity. Given a set $X$, we denote the closure of $X$
by $\overline{X}$.

@@

Recall that, in the case that $k=i\alpha$ for some real-valued $\alpha$,
the oscillatory Stokes equations \cref{eq:ostokes}
are known as the modified Stokes equations and are of particular
interest for their application to the analysis and numerical
simulation of unsteady flow
\cite{Pozrikidis1992,biros2002embedded,
  jiang2013second,ladyzhenskaya1969mathematical}.
The equations are well-studied in that setting and
integral representations which lead to second kind
integral equations have been developed. We review
some of the relevant results here, translating to
the present, oscillatory setting.

@@

\begin{equation}
  \cS [\bmu] (\xx) = \int_\Gamma \GG (\xx,\yy) \bmu(\yy)
  \, dS(\yy)
\end{equation}
and
\begin{equation}
\label{eq:dlformula}
  \cD [\bmu] (\xx) = \pv \int_\Gamma \left ( \TT_{\cdot,\cdot,\ell}(\bx,\by)
  \nu_\ell(\yy)
  \right )^\intercal \bmu(\yy) \, dS(\yy) \; ,
\end{equation}
where the \pv indicates that the integral is to be
evaluated in the principal value sense. 
For two vector valued
functions $\ff$ and $\bg$ defined on $\Gamma$, consider the bilinear
form

\begin{equation} \label{eq:bi_form}
  \langle \ff , \bg \rangle = \int_\Gamma \ff \cdot \bg dS \; .
\end{equation}
The definition of the adjoint used throughout the paper will be
the one induced by this form.
The adjoint of $\cD$ is of particular interest; we have
\begin{equation}
  \cD^{\intercal} [\bmu](\bx) = 
  \pv \int_\Gamma \left ( \TT_{\cdot,\cdot,\ell}(\bx,\by)\nu_\ell(\xx)
  \right ) \bmu(\yy) \, dS(\yy) \; .
\end{equation}

\begin{remark}
The operators $\bS,\bD,\cD$ and $\cD^{\intercal}$
depend on the Helmholtz parameter $k$ of the oscillatory Stokes equation.
In places where it is essential to highlight this dependence, in a slight
abuse of notation, we will use the symbols $\bS_{k}, \bD_{k},\cD_{k}$ and $\cDt_{k}$ to 
denote this dependence.
\end{remark}

@@

\begin{remark}
The operators $\bS,\bD, \cS, \cD$, and $\cW$  depend on the boundary 
of the region $\Gamma$. In places where it is necessary to highlight this dependence
we will use $A^{\Gamma}$ instead of the operator $A$.
\end{remark}

>>>>

\section{Mathematical Preliminaries}
\label{sec:prelim}
In this paper, vector-valued
quantities are denoted by bold, lower-case letters
(e.g. $\bh$), while tensor-valued quantities are bold
and upper-case (e.g. $\mathbf{T}$). 
Subscript indices of non-bold characters (e.g. $h_j$ or $T_{ij\ell}$)
are used to denote the entries within a vector ($\bh$) or tensor ($\bT$).
We use the standard Einstein summation convention; in other words, 
there is an implied sum taken over the repeated indices of 
any term (e.g. the symbol $a_{j} b_{j}$ is used to represent the sum
$\sum_{j} a_{j} b_{j}$).
If $\xx = (x_1,x_2)^\intercal$, then $\xx^\bot = (-x_2,x_1)^\intercal$.
Similarly, $\nabla^\bot = (-\partial_{x_2},\partial_{x_1})^\intercal$.
Upper-case script characters (e.g. $\mathcal{K}$) are reserved for
operators on Banach spaces, with $\mathcal{I}$ denoting the
identity. Given a set $X$, we denote the closure of $X$
by $\overline{X}$.

@@

Recall that, in the case that $k=i\alpha$ for some real-valued $\alpha$,
the oscillatory Stokes equations \cref{eq:ostokes}
are known as the modified Stokes equations and are of particular
interest for their application to the analysis and numerical
simulation of unsteady flow
\cite{Pozrikidis1992,biros2002embedded,
  jiang2013second,ladyzhenskaya1969mathematical}.
The equations are well-studied in that setting and
integral representations which lead to second kind
integral equations have been developed. We review
some of the relevant results here, translating to
the oscillatory setting.

@@

\begin{equation}
  \cS [\bmu] (\xx) = \int_\Gamma \GG (\xx,\yy) \bmu(\yy)
  \, dS(\yy)
\end{equation}
and
\begin{equation}
\label{eq:dlformula}
  \cD [\bmu] (\xx) = \pv \int_\Gamma \left ( \TT_{\cdot,\cdot,\ell}(\bx,\by)
  \nu_\ell(\yy)
  \right )^\intercal \bmu(\yy) \, dS(\yy) \; ,
\end{equation}
where the \pv indicates that the integral is to be
evaluated in the principal value sense.

For two vector valued
functions $\ff$ and $\bg$ defined on $\Gamma$, consider the bilinear
form
\begin{equation} \label{eq:bi_form}
  \langle \ff , \bg \rangle = \int_\Gamma \ff \cdot \bg dS \; .
\end{equation}
The definition of the adjoint used throughout the paper will be
the one induced by this form.

The adjoint of $\cD$ with respect to the above bilinear form
is of particular interest; and is given by
\begin{equation}
  \cD^{\intercal} [\bmu](\bx) = 
  \pv \int_\Gamma \left ( \TT_{\cdot,\cdot,\ell}(\bx,\by)\nu_\ell(\xx)
  \right ) \bmu(\yy) \, dS(\yy) \; .
\end{equation}

@@

\begin{remark}
The operators $\bS,\bD,\cD$ and $\cD^{\intercal}$
depend on the Helmholtz parameter $k$ of the oscillatory Stokes equation.
In places where it is essential to highlight this dependence, in a slight
abuse of notation, we will use the symbols $\bS_{k}, \bD_{k},\cD_{k}$ and $\cDt_{k}$ to 
denote this dependence.
Similarly, we will use $A^{\Gamma}$ instead of the operator $A$ to highlight 
the dependence of the operator $A$ on the boundary of the region $\Gamma$.
\end{remark}
